# Colours

## Color Brewer

* We've already used colours in some of the previous code examples. Particularly the palettes in the [RColorBrewer package](https://www.datanovia.com/en/blog/the-a-z-of-rcolorbrewer-palette).

* RColourBrewer contains ready made colour palettes we can view all at once like this.

```{r RColourBrewer}
# easy colours
RColorBrewer::display.brewer.all()
```

* We can also limit the display above to just those palettes that are friendly to people who are colour blind.

```{r}
display.brewer.all(colorblindFriendly = TRUE)
```

* We can create a palette of colours with a specific number of colours like this. And view them in a simple bar plot.

```{r}
mypalette <- RColorBrewer::brewer.pal(9,"Set2") 

graphics::barplot(1:9,col = mypalette)
```

## Prismatic

* [Prismatic](https://github.com/EmilHvitfeldt/prismatic) lets us play around with colour palettes by shifting the entire range of colours at once. For example, below we start with the terrain pallete in the prismatic package.

```{r}
terrain <- prismatic::colour(terrain.colors(10))
plot(terrain)
```

* Now we turn the palette grey.

```{r}
terrain_gray <- prismatic::clr_grayscale(terrain)
plot(terrain_gray)
```

* Darken the palette.

```{r}
terrain_darken <- prismatic::clr_darken(terrain)
plot(terrain_darken)
```

* Make the palette negative.

```{r}
terrain_negate <- prismatic::clr_negate(terrain)
plot(terrain_negate)
```

## Hand picked colours

* You can also hand pick your own colour palette from this [colour picking website](http://tristen.ca/hcl-picker/). It is recommended on a [detailed guid to ggplot colours](https://michaeltoth.me/a-detailed-guide-to-ggplot-colors.html) by Michael Toth.

* The [colour picking website](http://tristen.ca/hcl-picker/) was used to create and then copy and paste the colours in this palette below.

```{r colour_picker}
newpalette <- c("#804759","#713E4E","#633543","#562D39","#48252F","#3B1D25")

barplot(1:6,col = newpalette)
```