# Colours

## Color Brewer

* We've already used colours in some of the code. Particularly the pallettes in the [RColorBrewer package](https://www.datanovia.com/en/blog/the-a-z-of-rcolorbrewer-palette).

* It has some ready made palletes that we can view like this.

```{r RColourBrewer}
# easy colours
RColorBrewer::display.brewer.all()
```

* And the display can be limited to colour blind friendly palettes.

```{r}
display.brewer.all(colorblindFriendly = TRUE)
```

* we can set up a pallete of colours like this and view them in a simple bar plot

```{r}
mypalette <- RColorBrewer::brewer.pal(9,"Set2") 
graphics::barplot(1:9,col = mypalette)
```

## Prismatic

* [Prismatic](https://github.com/EmilHvitfeldt/prismatic) lets us play around with palettes by shifting the entire set at once. For example, starting with terrain..

```{r}
terrain <- prismatic::colour(terrain.colors(10))
plot(terrain)
```

* Turn the pallete grey

```{r}
terrain_gray <- prismatic::clr_grayscale(terrain)
plot(terrain_gray)
```

* Darken the pallete

```{r}
terrain_darken <- prismatic::clr_darken(terrain)
plot(terrain_darken)
```

* Make the pallete negative

```{r}
terrain_negate <- prismatic::clr_negate(terrain)
plot(terrain_negate)
```

## Hand picked colours

* You can also hand pick your own colours from this [colour picking website](http://tristen.ca/hcl-picker/) reccomedned on an [excellent blog post on using colours](https://michaeltoth.me/a-detailed-guide-to-ggplot-colors.html) in ggplot by by Michael Toth.

* The [colour picking website](http://tristen.ca/hcl-picker/) was used to create then copy and paste the colours below.

```{r colour_picker}
newpalette <- c("#804759","#713E4E","#633543","#562D39","#48252F","#3B1D25")
barplot(1:6,col = newpalette)
```
