# Colours

## RColorBrewer

* We have already been setting colours in some of the previous plots using  the [RColorBrewer package](https://www.datanovia.com/en/blog/the-a-z-of-rcolorbrewer-palette). RColourBrewer contains ready made colour palettes. We can view all the palletes it contains in onne picture like this.

```{r RColourBrewer}
# easy colours
RColorBrewer::display.brewer.all()
```

* We can also filter the picture above to show only palettes friendly to people who are colour blind.

```{r}
display.brewer.all(colorblindFriendly = TRUE)
```

* And we can create a palette of colours with a specific number of colours like this. Then view them in a simple bar plot.

```{r}
mypalette <- RColorBrewer::brewer.pal(9,"Set2") 

graphics::barplot(1:9,col = mypalette)
```

## Prismatic

* [Prismatic](https://github.com/EmilHvitfeldt/prismatic) lets us play around with colour palettes by shifting one aspect of all of the colours it contains at the same time. For example, below we start with the terrain pallete in the prismatic package.

```{r}
terrain <- prismatic::colour(terrain.colors(10))
plot(terrain)
```

* Now we turn the palette grey.

```{r}
terrain_gray <- prismatic::clr_grayscale(terrain)
plot(terrain_gray)
```

* Darken the palette.

```{r}
terrain_darken <- prismatic::clr_darken(terrain)
plot(terrain_darken)
```

* Make the palette negative.

```{r}
terrain_negate <- prismatic::clr_negate(terrain)
plot(terrain_negate)
```

## Hand picked colours

* You can also hand pick your own colour palette from this [colour picking website](http://tristen.ca/hcl-picker/). The website is recommended by Michael Toth in his excellent [detailed guide to ggplot colours](https://michaeltoth.me/a-detailed-guide-to-ggplot-colors.html).

* The [colour picking website](http://tristen.ca/hcl-picker/) was used to create and then copy and paste the colours into this palette below.

```{r colour_picker}
newpalette <- c("#804759","#713E4E","#633543","#562D39","#48252F","#3B1D25")

barplot(1:6,col = newpalette)
```